Compositional Quantitative Resource Analysis of C Programs

### 2014-06-07 paper/main.tex
catalog of examples in the paper

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include Speed [XXX], KoAT
[XXX], Pubs [XXX], RANK [XXX], and LOOPUS [XXX].
%
While the these tools can derive impressive results for
realistic software, there still exist several shortcomings that seem
to hamper the application of the existing techniques in practice.
\begin{enumerate}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant generation [XXX speed] or
  translation of the program into a term-rewriting system [costa,
  tacas] without providing \emph{verifiable certificates} for the
  correctness of the derived bound.
\end{enumerate}
%
While there has been progress there is still truth in the following
quote of Don Knuth from a recent interview [xxx].

\begin{quotation}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
\end{quotation}
%
%
In this paper, we present a novel cost analysis system for C programs
that works automatically, efficiently and precisely for many examples,
can be used interactively if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\begin{itemize}
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement that is parametric in a
  cost metric.  The semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost
  model.
\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional programs
  [XXX] and derives bounds that are non-negative linear combinations
  of sizes $|[x,y]|$ of intervals between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that the our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{itemize}
%
Our approach complements existing works since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource bound development
in which an automatically derived bound can be used in the
quantitative logic and a hand derived bound can be used by the
automatic analysis.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters.  To the best of our knowledge, our
system is the only one that can deal with resource like memory that
may become available during the execution.

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that
seem to be common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ and we have for example
$B(100,200,43) = 1043$.  
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce multiple results from the literature which were obtained
using methods based on abstract interpretation.  In contrast with
abstract interpretation--based methods, our technique does not require
any fixpoint computations on an abstract domain to obtain loop
invariants.  The mechanism we designed is able to leverage local
assertions such as $x < y$ that we collect along the branching points
to obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, mutually-recursive functions.

Following the development steps for automatic amortized analysis for
functional programs [XXX], we deliberately restrict our self to linear
bounds in the automatic analysis (there a no restrictions for the
manually-derived though).  More specifically, bounds have the form
$\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer variables
or constants.  The reason is mainly clarity of presentation, since
an extension to multivariate resource polynomials [XXX] would make the
inference rules considerably more complected and should better be
described separately.  However, we developed the linear inference
system so that the extension to polynomial bounds should work
smoothly.

@@

\subsection{Amortized Analysis}
The idea that drives the design of our system is amortized
analysis [XXX].  Say that a program $p$ evaluates on a heap $H$
by consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.
The basic idea of amortized analysis is to define a potential function
$\Phi$ from heaps to natural numbers, such that $\Phi(H) \ge n$.

@@

\bibliographystyle{abbrvnat}
\bibliography{lit}
\addcontentsline{toc}{section}{Bibliography}
\end{document}

>>>>

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include Speed [XXX], KoAT
[XXX], Pubs [XXX], RANK [XXX], and LOOPUS [XXX].
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of the existing techniques in practice.
\begin{enumerate}
\item Analysis tools are black boxes that either deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\item Existing techniques often rely on complex external tools such as
  abstract interpretation based invariant generation [XXX speed] or
  translation of the program into a term-rewriting system [costa,
  tacas] without providing \emph{verifiable certificates} for the
  correctness of the derived bound.
\end{enumerate}
%
While there has been lot of progress in static quantitative analysis,
Don Knuth correctly points out in a recent interview [XXX] that the
state-of-the-art in formal quantitative methods still falls short
in comparison with semantic techniques [xxx].
%
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-3ex}
  \begin{flushright}
    -- Donald Knuth, 2014
  \end{flushright}
\end{quote}
%
%
In this paper, we present a novel cost analysis system for C programs
that works automatically, efficiently and precisely for many examples,
can be used interactively if the automation fails, and produces
verifiable certificates for the bounds.  Our contributions are as
follows.
%
\item We define a operational cost semantics for C programs that
  accounts a constant cost for each statement which is parametric in a
  cost metric.  The semantics models cost of terminating and
  non-terminating executions.  A specific \emph{tick} function can be
  inserted in the source code to further customize the cost
  model.
\item We describe an automatic static analysis technique for deriving
  bounds for the cost that is defined by the operational semantics.
  It is based on amortized resource analysis for functional programs
  [XXX] and derives bounds that are non-negative linear combinations
  of sizes $|[x,y]|$ of intervals between program variables $x,y$.
\item We show with a publically available prototype implementation and
  an experimental evaluation that the our novel amortized resource
  analysis works efficiently for realistic programs and that the
  derived constant factors in the bounds are close to or identical
  with the optimal ones.
\item We define a quantitative Hoare logic for interactively deriving
  cost bounds for C programs.  The logic is implemented and proved
  sound in the Coq Proof Assistant with respect to CompCert C
  light.  We show that our automatic analysis readily produces
  derivations in the quantitative Hoare logic.
% \item We prove the completeness of the program logic
%   with respect to the operational cost semantics.
\end{itemize}
%
Our approach complements existing works since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  The quantitative Hoare logic and the automatic amortized
analysis enable semi-automatic interactive resource bound development
in which an automatically derived bound can be used in the
quantitative logic and a hand derived bound can be used by the
automatic analysis.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  To the best of our
knowledge, our system is the only one that can deal with resource like
memory that may become available during the execution.

The main innovation that makes this work practical is our new
automatic amortized analysis for C code.  It targets programs that are
common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]|$ of
intervals of integer variables.  A typical bound would be $B(x,y,z) =
10{\cdot}|[x,y]| + 1{\cdot}|[z,0]|$ and we have for example
$B(100,200,43) = 1043$.
%
Despite the apparent simplicity of the new analysis system, it is able
to reproduce multiple results from the literature which were obtained
using methods based on abstract interpretation.  In contrast with
abstract interpretation--based methods, our technique does not require
any fixpoint computations on an abstract domain to obtain loop
invariants.  The mechanism we designed is able to leverage local
assertions such as $x < y$ that we collect along the branching points
to obtain global invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program is
a resource usage bound for the C program.  We can naturally, handle
advanced control flow such as break, return, iteration over negative
counter variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs [XXX], we deliberately restrict our self to linear
bounds in the automatic analysis (there a no restrictions for the
manually-derived bounds in the logic).  More specifically, bounds have
the form $\sum_{a,b} q_{(a,b)} |[a,b]|$ where $a$ and $b$ are integer
variables or constants.  The reason is mainly clarity of presentation,
since an extension to multivariate resource polynomials [XXX] would
make the inference rules considerably more complected and should
better be described separately.  However, we developed the linear
inference system so that the extension to polynomial bounds shall
work smoothly.

We implemented our analysis for CompCert Clight, a subset of C in
which loops can only be exited with a \code{break} statement and
expressions are free of side effects.  C programs can for instance be
compiled to Clight programs using the verified CompCert C compiler.

Our new quantitative Hoare logic for CompCert Clight is a
generalization of a program logic that we recently developed to
formally very stack-space bounds in Coq [PLDI].  In a nutshell, we
replace the boolean predicates (pre and post conditions) of Hoare
logic with numeric predicates that map a program state to a
non-negative number or infinity.  The intuitive meaning of a
quantitative Hoare triple $\htriple{P}{C}{Q}$ that $C$ consumes $P(S)$
resources when executed in starting state $S$, and that $Q(S')$
resources are left if $C$ terminates in state $S'$.  Infinity
corresponds to \emph{false} in classic Hoare logic.  Our quantitative
logic is proved sound with respect to a cost semantic for Clight that
accounts a constant cost for each evaluation step.  This constant cost
is parametric in a resource metric that can be defined by the user.

@@

\subsection{Amortized Analysis}
The idea that drives the design of our system is amortized analysis
[XXX].  Say that a program $p$ evaluates on a program state $H$ by
consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.  The
basic idea of amortized analysis is to define a potential function
$\Phi$ that maps program states to non-negative numbers, such that
$\Phi(H) \ge n$.

@@

% \acks

% This research is based on work supported in part by NSF grants 1319671
% and 1065451, and DARPA grants FA8750-10-2-0254 and FA8750-12-2-0293.
% Any opinions, findings, and conclusions contained in this document are
% those of the authors and do not reflect the views of these agencies.

\appendix

\section{Catalog of Automatically Analyzed Programs}

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, we use cost metric that counts cost $1$ for comparisons
and assignments and cost $0$ for all other operations.  Of course, the
examples can also be analyzed with any other cost metric.

We assume that the variables in the code snippets are the inputs to
the program. Some of the examples contain constants on which the
computed bound depends.  These constants are randomly chosen to
present an example but the analysis works for other constants as well.
Note however that it is sometimes crucial that constants are positive
(or negative) and that the same constant at different places.

\begin{figure*}[t!]
\newlength{\progwidth}
\setlength{\progwidth}{.19\linewidth}
  \centering

  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x>0) {
    x=x-1;
    y=y+2;
  }
  while (y>0) {
    y=y-1;
  }
  while (y>0) {
    y=y+1;
  }
   \end{lstlisting}

$4|[0,x]| + [|0,y|]$
\\[.7\baselineskip]
      {\bf t07}
    \end{center}
  \end{minipage}%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (z-y>0) {
    y = y+1;
  }
  while (y>9) {
    y=y-10;
  }
   \end{lstlisting}

$1.1|[y,z]| + 0.1[|0,y|]$
\\[.7\baselineskip] 
      {\bf t08}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x-y>0) {
    if (*)
      y=y+1;
    else
      x=x-1;
  }
   \end{lstlisting}

$|[y,x]|$
\\[.7\baselineskip]
      {\bf t10}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x>0) {
    x=x-1;
    if (*) 
      y=y+1;
    else {
      while (y>0)
	y=y-1;
    }
  }
   \end{lstlisting}

$3|[0,x]| + |[0,y]|$
\\[.7\baselineskip]
      {\bf t13}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  n=n+1;
  y=y+1000;
  while (y>=100){
    y=y-100;
  }
}
   \end{lstlisting}

$12|[n,0]| + 0.1|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (x>y) {
    x=x-1;
    x=x+1000;
    y=y+1000;
  }
  while (y>0) {
    y=y-1;
  }
  while (x<0) {
    x=x+1;
  }
   \end{lstlisting}

$1004|[y,x]|+|[x,0]|+|[0,y]|$
\\[.7\baselineskip]
      {\bf t28}
    \end{center}
  \end{minipage}

   \caption{Amortization and Compositionality}
  \label{fig:cat1}
\end{figure*}

\paragraph{Amortization and Compositionality}

\pref{fig:cat1} shows code snippets that need amortization are
compositionality to obtain a whole program bound.

Example \emph{t07} demonstrates two different features of the
analysis.  For one thing it shows that we can precisely track size
changes inside loops.  In the first loop, we increment $y$ by $2$ in
each of the $|[0,x]|$ iterations.  For another thing it shows that we
automatically recognize dead code if we find conflicting assertions on
a branching path: After the second loop we know $y \leq 0$ and as a
result can assign arbitrary potential inside the third loop where we
know that $y>0$.  As a result, we obtain a tight bound.

Example \emph{t08} shows the ability of the analysis to handle
negative and non-negative numbers.  Note that there are no
restrictions on the signs of $z$ and $y$.  We also see again that we
accurately track the size change of $y$ in the first loop.
Furthermore, \emph{t08} shows that we do not handle the constants $1$
or $0$ in any special way.  In all examples you could replace $0$ and
$1$ with other constants like we did in the second loop and still
derive a tight bound.  The only information, that the analyzer needs
is $y \geq c$ before assigning $y = y - c$.

In example \emph{t10} we also do not restrict the inputs $x$ and $y$.
They can be negative, positive, or zero.  The star {\tt *} in the
conditional, stands for an arbitrary assertion.  In each branch of the
conditional we can obtain a constant potential $1$ since the interval
size $|[y,x]|$ is decreasing.

Example \emph{t13} shows how amortization can be used to handle tricky
nested loops.  The outer loop is iterated $[|0,x|]$ times.  In the
conditional, we either (the branching condition is again arbitrary)
increment the variable $y$ or we execute an inner loop in which $y$ is
counted back to $0$.  Then analysis computes a tight linear bound for
this program.  Again, the constants $0$ and $1$ in the inner loop can
as well be replace by something more interesting, say $9$ and $10$
like in example \emph{t08}.  Then we still obtain a tight linear
bound.

Example \emph{t27} is similar to example \emph{t13}.  Instead of
decrements the variable $x$ in the outer loop we this time increment
the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
we increment the variable $y$ by $1000$.  We then execute an inner
loop that increments $y$ by $100$ until $y=0$.  The analysis can
derive that only the first execution of the inner loop depends on the
initial value of $y$.  We again derive a tight bound.

Example \emph{t28} is particularly interesting.  In the first loop we
decrement the size $|[y,x]|$.  However, we also shift the interval
$[y,x]$ to the interval $[y+1000,x+1000]$.  The analysis can derive
that this does not change the size of the interval and computes the
tight loop bound $3|[y,x]|$.  The additional to loops are in the
program to show that the size tracking in the first loop works
accurately.  The second loop is executed $|[0,y]| + 1000|[y,x]|$ times
in the worst case.  The third loop is executed $|[x,0]| + |[y,x]|$ in
the worst case (if $x$ and $y$ are negative).

\bibliographystyle{abbrvnat}
\bibliography{lit}

\end{document}
